A meta analysis of tournaments and an evaluation of performance in the Iterated Prisoner's Dilemma.

### 2019-10-24 paper/main.tex
draft people hanging it to Vince

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
%%%%%%%%%%%%%%%%%

@@

@@

\title{A meta analysis of tournaments and an evaluation of performance in the
Iterated Prisoner's Dilemma.}
\author{Nikoleta E. Glynatsi, Dr Vincent A. Knight}
\date{}

@@

The Iterated Prisoner's Dilemma has been used for decades as a powerful model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. Most of these strategies are
now accessible due to an open source package; Axelrod-Python. This manuscript
make use of Axelrod-Python to conduct a meta analysis of 40000 Iterated
Prisoner's Dilemma tournaments. The aim is to evaluate the performance of
numerous strategies and finally answer the explore the factors of success in the
game.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a repeated two player game that models
situations in which self-interest clashes with collective interest. At each turn,
the players simultaneously and independently make a choice between cooperation (C) and
defection (D) whilst having memory of the prior interactions.
The payoffs at each given turn are defined by the matrix,

@@

where \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature~\cite{Axelrod1981}, and in this paper, are $R=3, P=1, T=5, S=0$.

Since the computer tournaments of R. Axelrod in 1980s several academic papers
are published in the field regarding the performance of strategies in the IPD.
In the 80's following the strong performance of Tit For Tat in both Axelrod's
computer tournaments~\cite{Axelrod1980a, Axelrod1980b}, and moreover in a series
of evolutionary experiments~\cite{Axelrod1981}, the strategy was thought as the
most robust basic strategy in the IPD. However, the strategy was shown to
perform poorly in environments with noise~\cite{Bendor1991, Donninger1986,
Molander1985, Hammerstein1984}. More robust strategy in such setting were
introduced and were the new protagonists, such as, Nice and
Forgiving~\cite{Bendor1991}, Pavlov~\cite{Nowak1993} and Generous Tit For
Tat~\cite{Nowak1992}.

The $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament took place
in 2004 with 233 entries. The winning strategies were based on a mechanism of
teams. A team from Southampton University took advantage of the fact that a
participant was allowed to submit multiple strategies. They submitted a total of
60 of strategies that could recognised each other and colluded to increase one
members score~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}.
Yet again, in 2012 another set of strategies was introduced as the dominant set
of strategies~\cite{Press2012}. These were called zero-determinant strategies,
and by forcing a linear relationship between the payoffs they can ensure that
they will never receive less than their opponents. However, in~\cite{Harper2017}
a tournament containing over 200 strategies, zero-determinant, was performed and
none of the zero-determinant strategies ranked in top spots. Instead, the top
ranked strategies were a set of evolved strategies based lookup
tables~\cite{Axelrod1987}, hidden markov models~\cite{Harper2017} and finite
state automata~\cite{Miller1996}.
%TODO reference the archetypes of the archetypes used the IPD? Did IPD for now.

Thus, the following question is raised here: which are the true dominant
strategies in the iterated prisoner's dilemma?
This manuscript uses the open source package
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments using as many strategies as possible from the literature. The aim is
to evaluate the performance of these strategies in a tournament and futhermore,
explore the factors of their success. This is done not for standard round robin
tournaments, but also for noisy, probabilistic ending and noisy probabilistic
ending tournaments.

The different tournaments and the data generating process are covered in
Section~\ref{section:data_collection}. Section~\ref{section:top_performances},
covers the best performed strategies for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance and finally in Section~\ref{section:discussion}
the results are discussed and concluded in Section~\ref{section:conclusion}.

\section{Data generating process}\label{section:data_collection}

For the purposes of this manuscript a data set containing results on IPD tournaments
has been generated and is available at. This
was done using the open source package Axelrod-Python~\cite{axelrodproject},
more specifically, version 3.0.0. Axelrod-Python allows for different types of
IPD computer tournaments to be simulated whilst
containing a list of over 180 strategies. Most of these are strategies described
in the literature with a few exceptions being strategies that have been
contributed specifically to the package. Though Axelrod-Python features several
tournament types, this work considers only standard, noisy, probabilistic ending
and noisy probabilistic ending tournaments.

\textbf{Standard tournaments}, are tournaments similar to that of Axelrod's
in~\cite{Axelrod1980a}. There are \(N\) strategies which all play an iterated
game of \(n\) number of turns against each other. Note that self interactions
and a match against a random strategy are not included. Similarly, \textbf{noisy
tournaments} have \(N\) strategies and \(n\) number of turns but at each turn
there is a probability \(p\) that a player's action will be flipped.
\textbf{Probabilistic ending tournaments}, are of size \(N\) and after each turn
a match between strategies ends with a given probability \(e\). Finally,
\textbf{noisy probabilistic ending} tournaments have both a noise probability
\(p\) and an ending probability \(e\). For smoothing the simulated results a
tournament is repeated for \(k\) number of times. The winner of each tournament
is based on the average score a strategy achieved and not by number of wins.

%  A summary of
% each tournaments' parameters is given in Table~\ref{table:tournaments_parameters}.

@@

The process of generating data implemented in this manuscript is given by
Algorithm~\ref{algorithm:data_generation}. For each trial a random
size \(N\) is selected, and from the list of 186 strategies
in~\cite{axelrodproject}, a random list of \(N\) strategies is chosen. The 186
strategies used here are given in the Appendix~\ref{app:list_of_players}. For the
given list of strategies a standard, a noisy, a probabilistic ending and a noisy
probabilistic ending tournament are performed and repeated \(k\) times.
The parameters for the tournaments as well as the number of repetitions are
selected once for each trial. The parameters and their respective minimum and
maximum values are given by Table~\ref{table:parameters_values}.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.6\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    parameter & parameter explanation &   min value & max value \\
    \midrule
    $N$ & number of strategies  & 3 & 195 \\
    $k$ & number of repetitions  & 10 & 100 \\
    $n$ & number of turns      & 1 & 200 \\
    $p$ & probability of flipping action at each turn  & 0 & 1   \\
    $e$ & probability of match ending in the next turn & 0 & 1   \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Data generation parameters' values}
    \label{table:parameters_values}
\end{table}

The source code for the data generating process as well as the source code for
the analysis which will be discussed in the following sections have been written
following best practices~\cite{Aberdour2007, Benureau2018}. It has been packaged
and is available here.

\begin{algorithm}[!htbp]
    \setstretch{1.35}
    \ForEach{\text{seed} $\in [0, 12285]$}{
        $N \gets \text{randomly select integer}\in [N_{min}, N_{max}]$\;
        $\text{players} \gets  \text{randomly select $N$ players}$\;
        $k \gets  \text{randomly select integer}\in [k_{min}, k_{max}]$\;
        $n \gets  \text{randomly select integer}\in [n_{min}, n_{max}]$\;
        $p \gets  \text{randomly select float}\in [p_{min}, p_{max}]$\;
        $e \gets   \text{randomly select float}\in [e_{min}, e_{max}]$\;
        \vspace{0.4cm}
        $\text{result standard}$ $\gets$ Axelrod.tournament$(\text{players}, n, k)$\;
        $\text{result noisy}$ $\gets$ Axelrod.tournament$(\text{players}, n, p, k)$\;
        $\text{result probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, e, k)$\;
        $\text{result noisy probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, p, e, k)$\;

    }
    \KwRet{result standard, result noisy, result probabilistic ending,
    result noisy probabilistic ending}\;
    \caption{Data generating Algorithm}
    \label{algorithm:data_generation}
\end{algorithm}

A total of 12,285 trials of Algorithm~\ref{algorithm:data_generation} have been
performed. For each trial the results for 4 different tournaments were collected,
thus a total of 49,140 $(12,285 \times 4)$ tournament results have been
retrieved. Each tournament outputs a result summary in the form of
Table~\ref{table:output_result}.

The result summary has a length \(N\) because each row contains information for
each strategy that participated in the tournament. The information include the
strategy's rank, median score, the rate with which the strategy cooperated
$(C_r)$, it's wins and the probability that the strategy cooperated in the
opening move. Moreover, the rates of a strategy being in any of the four states
($CC, CD, DC, DD$), and the rate of which the strategy cooperated after each
state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[!htbp]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result.}\label{table:output_result}
\end{table}

The \textbf{normalised rank} is a measure which was manually included. The normalised
rank, denoted as $r$, is calculated as a strategy's rank divided by the
tournament's size ($N$). The normalised rank will be used in the next section to
evaluate the performance of strategies.

@@

This section evaluates the performance of 186 strategies. The performance of
each strategy will be evaluated for each type of tournament independently,
followed by an evaluation of their performance over all the simulated
tournaments of this work. Each strategy could have participated in multiple
tournaments of the same type (on average each participated in 5690 different
tournaments). For example Tit For Tat has participated in a total of 5569
tournaments of each type. The strategy's normalised rank distribution in these
is given in Figure~\ref{fig:tit_for_tat_r_distribution}. As a result, of the
multiple entries of strategies their performance is evaluated based on the
\textbf{median normalised rank} denoted as \(\bar{r}\). A value of \(\bar{r} =
0\) corresponds to a strategy winning the tournament where a a value of
\(\bar{r} = 1\) corresponds to the strategy coming last.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.8\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a normalised
    rank of 0.339.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

The result of each tournament type are based on $12,285$ trials. The strategies
are ranked based on the median normalised rank they achieved and the top 15
strategies of each tournament are given in Table~\ref{table:top_performances}.

@@

In standard tournaments 10 of the 15 top strategies are strategies introduced
in~\cite{Harper2017}. These have been trained using reinforcement learning
algorithms and are based on finite state automata (FSM), hidden markov models
(HMM), artificial neural networks (ANN), lookup tables (LookerUp) and stochastic
lookup tables (Gambler). Specifically, the have been trained against the
strategy list of~\cite{axelrodproject} in standard tournaments. Thus their
performance is to be expected. DoubleCrosser, and Fool Me Once, are not from the
literature but from~\cite{axelrodproject}. DoubleCrosser is a strategy that
makes use of the length of the match because is set to defect on the last two
rounds. The strategy is expected to not perform as well in probabilistic ending
tournaments. Finally, Winner 12~\cite{mathieu2017} and DBS are both from the the
literature. DBS~\cite{Au2006} is strategy specifically designed for noisy
environments, however, it ranks highly only in standard ones.

Figure~\ref{fig:std_results} gives the distributions of $\bar{r}$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median is at 0.075. This indicates that the top ranked strategies are dominating
strategies in standard tournaments. They are very likely to perform well in
standard tournament despite the number of opponents, the opponents, the turns
etc. This does not hold for all the tournament types as it will be discussed
in the later parts.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_standard.pdf}
    \caption{$\bar{r}$ distributions of top 15 strategies in standard tournaments.}\label{fig:std_results}
\end{figure}

The top strategies in noisy tournaments include the two strategies, Tit For 2
Tats~\cite{Axelrod1980b} and Hard Tit For 2 Tats~\cite{Stewart2012}, which are
strategies that will defect only after the have received two defections from the
opponent. The Retaliate strategies are set of strategies
from~\cite{axelrodproject} that start by cooperating but will retaliate once the
opponent's wins and defections surpass a curtain threshold.
ShortMem~\cite{Carvalho2013}, Grumpy, $e$ and $\phi$ are strategies that make
decisions based on the cooperations to defections ratio. In $5^{\text{th}}$ and
$6^{\text{th}}$ place are the strategies Cycler Hunter and  Risky QLearn. Cycler
Hunter tries to extort strategies that play cyclically and Risky QLearn uses a Q
learning algorithm. Notably, a deterministic and one of the most simple
strategies in game is ranked $3^{\text{rd}}$. That is Cooperator, a strategy
that just cooperates.

From Figure~\ref{fig:std_results} it is evident that the normalised rank distributions
in noisy environments are more variant and have higher median values compared
to standard tournaments. The distributions are skewed both towards
0 and 1 which indicates that though the top ranked strategies mainly
performed well (medians $< 0.3$) there are several tournaments that they performed
worse than the 60\% of the participants.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.7\textwidth]{../images/performance_noise.pdf}
    \caption{\(\bar{r}\) distributions for best performed strategies in noisy tournaments.}
    \label{fig:noisy_results}
\end{figure}

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014} which are strategies
based on finite state automata introduced by Daniel and Wendy Ashlock. These
strategies have been evolved using reinformed learning, however, there were
trained to maximise their payoffs in tournaments with fixed turns (150
specifically) and not in probabilistic ending ones.
In probabilistic ending tournaments it appears that the top ranks are mostly
occupied by defecting strategies which include Better and Better, Gradual
Killer, Hard Prober (all from ~\cite{prison}), Bully (Reverse Tit For
Tat)~\cite{Nachbar1992} and Defector. Thus, it's surprisingly that EasyGo and
Fool Me Forever are ranked $14^{\text{th}}$ and $15^{\text{th}}$. These
strategies are actually the same; they will defect until their opponent defect,
then they will cooperate until the end. Both strategies have repeatedly ranked highly
as shown in Figure~\ref{fig:probend_results} and there are cases for which they
were the winners of the tournament.

>>>>

%%%%Packages%%%%
\usepackage{amsmath}
\usepackage{booktabs}
\usepackage{graphics}
\usepackage{multicol}
\usepackage[ruled,vlined]{algorithm2e}
\usepackage{setspace}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{color,colortbl}
\usepackage{array}
\usepackage{booktabs}
\usepackage{tabularx}
\usepackage{wrapfig, blindtext}
\usepackage[table]{xcolor}
%%%%%%%%%%%%%%%%%

@@

\newcommand{\uniquenumberofseeds}{\input{unique_number_of_seeds.tex}}
\newcommand{\numberofalltournaments}{\input{number_of_all_tournaments.tex}}

@@

\title{A meta analysis of tournaments and an evaluation of performance in the
Iterated Prisoner's Dilemma.}
\author{Nikoleta E. Glynatsi, Vincent A. Knight}
\date{}

@@

The Iterated Prisoner's Dilemma has been used for decades as a model of
behavioural interactions. From the celebrated performance of Tit for Tat, to the
introduction of the zero-determinant strategies, to the use of sophisticated
structures such as neural networks, the literature has been exploring the
performance of strategies in the game for years. Most of these strategies are
now accessible due to an open source package; Axelrod-Python. This manuscript
make use of Axelrod-Python to conduct a meta analysis of Iterated Prisoner's
Dilemma tournaments. The aim is to evaluate the performance of 194 strategies
over \numberofalltournaments Iterated Prisoner's Dilemma tournaments and to
explore the factors of their success.
\end{abstract}

@@

The Iterated Prisoner's Dilemma (IPD) is a repeated two player game that models
behavioural interactions, and more specifically, interactions where
self-interest clashes with collective interest. At each turn of the game both
players, simultaneously and independently, decide between cooperation (C) and
defection (D) whilst having memory of their prior interactions. The payoffs for each
player, at each turn, is influenced by their own choice and the choice of the
other player. The payoffs of the game are generally defined by:

@@

where \(T > R > P > S\) and \(2R > T + S\). The most common values used in
the literature~\cite{Axelrod1981} are $R=3, P=1, T=5, S=0$. These values are also
used in this work.

Conceptualising strategies and understanding the best way of playing the game
has been of interest to the scientific community since the formulation of the
game in 1950~\cite{Flood1958}. Following the computer tournaments of Axelrod in
1980's~\cite{Axelrod1980a, Axelrod1980b}, a strategy's performance in a round
robin computer tournament became a common evaluation technique for newly designed
strategies. Today more than 200 strategies exist in the literature and several
tournaments, excluding Axelrod's, have been undertaken~\cite{Bendor1991,
Harper2017, Kendall2007, Stephens2002, Stewart2012}.

In the 80's, Axelrod performed two computer tournaments~\cite{Axelrod1980a,
Axelrod1980b}. The contestants were strategies submitted in the form of computer
code. They competed against all other entries, a copy of themselves and a random
strategy. The winner was decided on the average score a strategy achieved. The
winner of both tournaments was the simple strategy Tit For Tat which cooperated
on the first turn and then simply copied the previous action of it's opponent.
Due to the strategy's strong performance in both tournaments, and moreover in a
series of evolutionary experiments~\cite{Axelrod1981}, Tit For Tat was thought
to be the most robust basic strategy in the IPD.

However, further research proved that the strategy had weakness, and more
specifically, it was shown that the strategy suffered in environments with
noise~\cite{Bendor1991, Donninger1986, Molander1985, Hammerstein1984}. This was
mainly due to the strategy's lack of generosity and contrition. The strategy was
quick to punish a defection, and in a noisy environment it could lead to a
repeated cycle of defections and cooperations. Some new strategies, more
robust in tournaments with noise, were soon introduced and became the new
protagonists of the game. These include Nice and Forgiving~\cite{Bendor1991},
Pavlov~\cite{Nowak1993} and Generous Tit For Tat~\cite{Nowak1992}.

In 2004, a $20^{\text{th}}$ Anniversary Iterated Prisoner Dilemma Tournament
took place with 233 entries. This time the winning strategy was not designed on
a reciprocity based approach but on a mechanism of
teams~\cite{J.P.Delahaye1993Lp, J.P.Delahaye1995LIeP, A.Rogers2007Ctpw}. A team
from Southampton University took advantage of the fact that a participant was
allowed to submit multiple strategies. They submitted a total of 60 strategies
that could recognised each other and colluded to increase one members score.
This resulted with three of the strategies to be ranked in the top spots. The
performance of the Southampton University team received mixed attention, though
they had won the tournament as stated in~\cite{us_blog} "technically this
strategy violates the spirit of the Prisoner's Dilemma, which assumes that the
two prisoners cannot communicate with one another".

Another set of IPD strategies that have received a lot of attention, the
American Mathematical Society's news section stated that ``the world of game
theory is currently on fire'', are the zero-determinants~\cite{Press2012}. By
forcing a linear relationship between the payoffs the zero-determinant
strategies can ensure that they will never receive less than their opponents.
Zero-determinant strategies are indeed a set of mathematically unique strategies
and robust in pairwise interactions, however, their simplicity and extortionate
behaviour have been tested. In~\cite{Harper2017} a tournament containing over
200 strategies, including zero-determinants, was ran. None of the zero-determinant
strategies ranked in top spots. Instead, the top ranked strategies were a set of
trained strategies based lookup tables~\cite{Axelrod1987}, hidden markov
models~\cite{Harper2017} and finite state automata~\cite{Miller1996}.

Though only a few selected pieces of work have been discussed, the IPD is
rich, and new strategies and competitions are being published every year.
The question, however, still remains the same: what is the best way to play the
game?

This manuscript uses an open source package called
Axelrod-Python~\cite{axelrodproject} to simulate a large number of computer
tournaments, using as many strategies as possible from the literature. The aim
is to evaluate the performance of 194 strategies in IPD tournaments, and
futhermore, explore the factors of their success. This is done not only for
standard round robin tournaments, but also for tournaments with noisy and
tournaments with a probabilistic ending. The different tournament types as well
as the data collection are covered in Section~\ref{section:data_collection}.
Section~\ref{section:top_performances}, focuses on the best performed strategies
for each type of tournament and overall.
Section~\ref{section:evaluation_of_performance}, explores the traits which
contribute to good performance, and finally the results are summarised in
Section~\ref{section:conclusion}.

This manuscripts uses several parameters. These are introduced in the following
sections, however, the full set of parameters and their definitions are given in
Appendix~\ref{app:parameters}. %TODO: fix table

\section{Data collection}\label{section:data_collection}

For the purposes of this manuscript a data set containing results of IPD tournaments
has been generated and is available at~\cite{data}. This
was done using the open source package Axelrod-Python~\cite{axelrodproject}, and
more specifically, version 3.0.0. Axelrod-Python allows for different types of
IPD computer tournaments to be simulated whilst
containing a list of over 180 strategies. Most of these are strategies described
in the literature with a few exceptions being strategies that have been
contributed specifically to the package. This paper make use of 194 strategies
implemented in version 3.0.0. A list of the strategies is given in the
Appendix~\ref{app:list_of_players}. Though Axelrod-Python features several
tournament types, this work considers only standard, noisy, probabilistic ending
and noisy probabilistic ending tournaments.

\textbf{Standard tournaments}, are tournaments similar to that of Axelrod's
in~\cite{Axelrod1980a}. There are \(N\) strategies which all play an iterated
game of \(n\) number of turns against each other. Note that self interactions
and a match against a random strategy are not included. Similarly, \textbf{noisy
tournaments} have \(N\) strategies and \(n\) number of turns, but at each turn
there is a probability \(p\) that a player's action will be flipped.
\textbf{Probabilistic ending tournaments}, are of size \(N\) and after each turn
a match between strategies ends with a given probability \(e\). Finally,
\textbf{noisy probabilistic ending} tournaments have both a noise probability
\(p\) and an ending probability \(e\). For smoothing the simulated results a
tournament is repeated for \(k\) number of times and the winner of each tournament
is based on the average score a strategy achieved and not by number of wins.

% A summary of each tournaments' parameters is given in
% Table~\ref{table:tournaments_parameters}.

@@

The process of collecting tournament results implemented in this manuscript is described by
Algorithm~\ref{algorithm:data_generation}. For each trial a random size \(N\) is
selected, and from the 194 strategies a random list of \(N\) strategies is
chosen. For the given list of strategies a standard, a noisy, a probabilistic
ending and a noisy probabilistic ending tournament are performed and repeated
\(k\) times. The parameters for the tournaments as well as the number of
repetitions are selected once for each trial. The parameters and their
respective minimum and maximum values are given by
Table~\ref{table:parameters_values}.

\begin{table}[!htbp]
    \begin{center}
        \resizebox{.6\textwidth}{!}{
        \begin{tabular}{lcccc}
    \toprule
    parameter & parameter explanation &   min value & max value \\
    \midrule
    $N$ & number of strategies  & 3 & 195 \\
    $k$ & number of repetitions  & 10 & 100 \\
    $n$ & number of turns      & 1 & 200 \\
    $p$ & probability of flipping action at each turn  & 0 & 1   \\
    $e$ & probability of match ending in the next turn & 0 & 1   \\
    \bottomrule
        \end{tabular}}
    \end{center}
    \caption{Data collection; parameters' values}
    \label{table:parameters_values}
\end{table}

The source code for the data collection as well as the source code for
the analysis, which will be discussed in the following sections, have been written
following best practices~\cite{Aberdour2007, Benureau2018}. It has been packaged
and is available here. %TODO archive code

\begin{algorithm}[!htbp]
    \setstretch{1.35}
    \ForEach{\text{seed} $\in [0, 11420]$}{
        $N \gets \text{randomly select integer}\in [N_{min}, N_{max}]$\;
        $\text{players} \gets  \text{randomly select $N$ players}$\;
        $k \gets  \text{randomly select integer}\in [k_{min}, k_{max}]$\;
        $n \gets  \text{randomly select integer}\in [n_{min}, n_{max}]$\;
        $p \gets  \text{randomly select float}\in [p_{min}, p_{max}]$\;
        $e \gets   \text{randomly select float}\in [e_{min}, e_{max}]$\;
        \vspace{0.4cm}
        $\text{result standard}$ $\gets$ Axelrod.tournament$(\text{players}, n, k)$\;
        $\text{result noisy}$ $\gets$ Axelrod.tournament$(\text{players}, n, p, k)$\;
        $\text{result probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, e, k)$\;
        $\text{result noisy probabilistic ending}$ $\gets$ Axelrod.tournament$(\text{players}, p, e, k)$\;

    }
    \KwRet{result standard, result noisy, result probabilistic ending,
    result noisy probabilistic ending}\;
    \caption{Data collection Algorithm}
    \label{algorithm:data_generation}
\end{algorithm}

A total of \uniquenumberofseeds trials of Algorithm~\ref{algorithm:data_generation} have been
run. For each trial the results for 4 different tournaments were collected,
thus a total of \numberofalltournaments $(\uniquenumberofseeds \times 4)$ tournament results have been
retrieved. Each tournament outputs a result summary in the form of
Table~\ref{table:output_result}.

The result summary, Table~\ref{table:output_result}, has \(N\) number of rows
because each row contains information for each strategy that participated in the
tournament. The information include the strategy's rank, median score, the rate
with which the strategy cooperated $(C_r)$, it's wins and the probability that
the strategy cooperated in the opening move. Moreover, the rates of a strategy
being in any of the four states ($CC, CD, DC, DD$), and the rate of which the
strategy cooperated after each state.

\newcolumntype{g}{>{\columncolor{Gray}}c}
\begin{table}[h!]
    \begin{center}
    \resizebox{\textwidth}{!}{
    \begin{tabular}{ccccccgcgcgcgcg}
    \toprule 
    & & & & & &   \multicolumn{8}{g}{Rates}  \\
    Rank & Name & Median score & Cooperation rating $(C_r)$ & Win & Initial C &
    CC & CD & DC & DD & CC to C & CD to C & DC to C & DD to C \\
    0 &  EvolvedLookerUp2 2 2 & 2.97 & 0.705 & 28.0 & 1.0 & 0.639 & 0.066 & 0.189 &
    0.106 & 0.836 & 0.481 & 0.568 & 0.8 \\
    1 &  Evolved FSM 16 Noise 05 & 2.875 & 0.697 & 21.0 & 1.0 & 0.676 & 
    0.020 & 0.135 & 0.168 & 0.985 & 0.571 & 0.392 & 0.07 \\
    2 & PSO Gambler 1 1 1 & 2.874 & 0.684 &  23.0 &     1.0 &    0.651 &    0.034 &    0.152 &    0.164
    & 1.000 & 0.283 & 0.000 & 0.136 \\
    3 &  PSO Gambler Mem1 &  2.861 &        0.706 &  23.0 &      1.0 &    0.663
    &    0.042 &    0.145 &    0.150 &  1.000 &  0.510 &  0.000 &  0.122 \\
    4 &          Winner12 &  2.835 &        0.682 &  20.0 &      1.0 &
    0.651 &    0.031 &    0.141 &    0.177 &  1.000 &  0.441 &  0.000 &  0.462 \\
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ &
    $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ & $\dots$ \\
    \bottomrule
    \end{tabular}}
\end{center}
\caption{Output result of a signle tournament.}\label{table:output_result}
\end{table}

A measure that has been manually included
is the \textbf{normalised rank}. The normalised rank, denoted as $r$, is
calculated as a strategy's rank divided by the tournament's size ($N$). The
normalised rank will be used in the next section to evaluate the performance of
strategies.

@@

This section evaluates the performance of 194 IPD strategies. The performance of
each strategy is evaluated in four tournament types, which were presented in Section
\ref{section:data_collection}, followed by an evaluation of their performance
over all the \numberofalltournaments simulated tournaments of this work.

Each strategy participated in multiple tournaments of the same type (on average 5180).
For example Tit For Tat has participated in a total of 5114
tournaments of each type. The strategy's normalised rank distribution in these
is given in Figure~\ref{fig:tit_for_tat_r_distribution}. Because of the strategies'
multiple entries their performance is evaluated based on the
\textbf{median normalised rank} denoted as \(\bar{r}\). A value of \(\bar{r} =
0\) corresponds to a strategy winning the tournament where a a value of
\(\bar{r} = 1\) corresponds to the strategy coming last.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.45\textwidth]{../images/tit_for_tat_r_distributions.pdf}
    \caption{Tit For Tat's $r$ distribution in tournaments. The best performance
    of the strategy has been in standard tournaments where it achieved a $\bar{r}$
    of 0.339.}
    \label{fig:tit_for_tat_r_distribution}
\end{figure}

The top 15 strategies for each tournament type based on \(\bar{r}\) are given in
Table~\ref{table:top_performances}.

@@

In standard tournaments 10 out of the 15 top strategies are introduced
in~\cite{Harper2017}. These are strategies based on finite state automata (FSM),
hidden markov models (HMM), artificial neural networks (ANN), lookup tables
(LookerUp) and stochastic lookup tables (Gambler) that have been trained using
reinforcement learning algorithms (evolutionary and particle swarm algorithms).
They have been trained to perform well against the strategies
in~\cite{axelrodproject} in a standard setting, thus their performance in the
specific setting was anticipated. DoubleCrosser, and Fool Me Once, are
strategies not from the literature but from~\cite{axelrodproject}. DoubleCrosser
is a strategy that makes use of the number of turns because is set to defect on
the last two rounds. The strategy was expected to not perform as well in
tournaments where the number of turns is not specified, but the strategy did not
perform well in tournaments with noise either. Finally, Winner
12~\cite{mathieu2017} and DBS~\cite{Au2006} are both from the the literature.
DBS is strategy specifically designed for noisy environments, however, it ranks
highly only in standard ones.

Figure~\ref{fig:std_results} gives the distributions of $r$ for the top
ranked strategies. The distributions are skewed towards zero and the highest
median is at 0.075. This indicates that the top ranked strategies are very
likely to perform well in a standard tournament despite the tournament size,
the opponents, the turns etc.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.55\textwidth]{../images/performance_standard.pdf}
    \caption{$r$ distributions of top 15 strategies in standard tournaments.}\label{fig:std_results}
\end{figure}

The top strategies in noisy tournaments include deterministic strategies, such
as Tit For 2 Tats~\cite{Axelrod1980b}, Hard Tit For 2 Tats~\cite{Stewart2012}
and Cycler CCCCCD, and strategies which decide their actions based on the
cooperations to defections ratio, such as ShortMem~\cite{Carvalho2013}, Grumpy,
$e$, $\pi$ and $\phi$ (all from~\cite{axelrodproject}). The Retaliate and
Limited Retaliate strategies are implemented in~\cite{axelrodproject} by the
same contributor. The strategies are designed to defect if the opponent has
tricked the strategy more often than \(x\%\) of the times that the strategy has
done the same. Finally, in $4^{\text{th}}$ and $9^{\text{th}}$ are Hunter
strategies which trying to extort, equivalently, strategies that play cyclically
and defectors.

From Figure~\ref{fig:noisy_results}, it is evident that the normalised rank
distributions in noisy environments are more variant with higher medians
compared to standard tournaments. The distributions are skewed both towards 0
and 1. This indicates that although the top ranked strategies mainly performed
well, there are several tournaments that they ranked in the bottom half.

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=.55\textwidth]{../images/performance_noise.pdf}
    \caption{\(r\) distributions for best performed strategies in noisy tournaments.}
    \label{fig:noisy_results}
\end{figure}

The 15 top ranked strategies in probabilistic ending tournaments include
Fortress 3, Fortress 4 (both introduced in~\cite{Ashlock2006}),
Raider~\cite{Ashlock2014} and Solution B1~\cite{Ashlock2014}, which are
strategies based on finite state automata introduced by Daniel and Wendy
Ashlock. These strategies have been evolved using reinformed learning, however,
there were trained to maximise their payoffs in tournaments with fixed turns
(150 specifically) and not in probabilistic ending ones. In probabilistic ending
tournaments it appears that the top ranks are mostly occupied by defecting
strategies. These include Better and Better, Gradual Killer, Hard Prober (all
from ~\cite{prison}), Bully (Reverse Tit For Tat)~\cite{Nachbar1992} and
Defector. Thus, it's surprisingly that EasyGo and Fool Me Forever are ranked
$14^{\text{th}}$ and $15^{\text{th}}$. These strategies are actually the same;
they will defect until their opponent defect, then they will cooperate until the
end. Both strategies have repeatedly ranked highly as shown in
Figure~\ref{fig:probend_results}, and there are cases for which they were the
winners of the tournament.

\begin{figure}[h!]
    \centering
    \includegraphics[width=.55\textwidth]{../images/performance_probend.pdf}
    \caption{\(r\) distributions for best performed strategies in probabilistic ending tournaments.}
    \label{fig:probend_results}
\end{figure}
